---
title: "Assignment"
author: "Eklak"
date: "2025-05-15"
output: html_document
---

# A. Install Packages
```{r Install Packages, include=FALSE}
install.packages("matlib")
install.packages("rsample")
install.packages("glmnet")
```


# B. Import Libraries required for analysis
```{r Load Libraries, include=FALSE}
library(matlib)
library(ggplot2)
library(rsample)
library(glmnet)
```


# C. Importing and Loading data from the CSV file
```{r Load dataset from the CSV, echo=TRUE}
rawData <- read.csv("data/dataset.csv")
head(rawData)
```




# C.1. Data Assignment to X and displaying it
```{r Creating Matrix X and Displaying data}
X <- as.matrix(rawData[, 1:4]) # Select the first four columns
# x1 -> Temperature (T)
# x3 -> Ambient Pressure (AP)
# x4 -> Relative Humidity (RH)
# x5 -> Exhaust Vacuum (V)
colnames(X)<-c("x1","x3","x4","x5")

X # Displaying the data
```


# C.2. Data Assignment to Y and displaying it
```{r Creating Matrix Y with x2 data and display it}
# x2 -> Net hourly energy output (EP)
Y <- as.matrix(rawData[, 5])   # Select the fifth x2 column (last one)
colnames(Y)<-c("y")

Y
```


# Task 1: Preliminary Data Analysis
## Plotting Time series data
```{r}
# Accessing input and output variables from matrix
x1 <- X[,1]
x3 <- X[,2]
x4 <- X[,3]
x5 <- X[,4]
y <- Y[,1]
```

```{r}
# Convert the input variables to time series objects using row index as time
x1_ts <- ts(x1)
x3_ts <- ts(x3)
x4_ts <- ts(x4)
x5_ts <- ts(x5)

y_ts <- ts(y)
```


## Plotting time series of input and target variables
```{r}
# Set up the plotting area to have 4 plots in one column with adjusted margins
par(mfrow = c(4, 1), mar = c(4, 4, 2, 1))

# Plot the time series of x1
plot(x1_ts, type = "l", col = "red", xlab = "Time", ylab = "x1", main = "Time Series of Input Signal: x1")

# Plot the time series of x3
plot(x3_ts, type = "l", col = "green", xlab = "Time", ylab = "x3", main = "Time Series of Input Signal: x3")

# Plot the time series of x4
plot(x4_ts, type = "l", col = "blue", xlab = "Time", ylab = "x4", main = "Time Series of Input Signal: x4")

# Plot the time series of x5
plot(x5_ts, type = "l", col = "purple", xlab = "Time", ylab = "x5", main = "Time Series of Input Signal: x5")


# Reset the plotting area to have 1 plot for the output variable
par(mfrow = c(1, 1))

# Plot the time series of x2
plot(y_ts, type = "l", col = "orange", xlab = "Time", ylab = "x2", main = "Time Series of Output Signal")


```

# Task 1.2 : Plotting distribution of each input CCPP signal
# Creating a density of all input signal X 

```{r}
density_of_X=density(X)
plot(density_of_X,main = "Combined Density plot of input signal X", col="brown")
```

## Creating histogram of X signal
```{r}
hist(X,freq = FALSE,main = "Histogram of Input signal X")
```

## Creating histogram of Y signal
```{r}
hist(Y,freq = FALSE,main = "Histogram of Output signal Y")
```

## Combining Histogram of X signal with density plot
```{r}
hist(X,freq = FALSE,main = "Density and Histogram plot of Input Signal X")
lines(density_of_X,lwd=2,col="brown")
rug(jitter(X))
```


## Histogram and density plot of individual input signal X and output signal y
```{r}
# Plot density for x1
x1_density <- density(x1)
hist(x1,freq = FALSE,main = "Histogram and density plot of x1",xlab = "x1 Signal")
lines(x1_density,lwd=2,col="red")
# Add the data-points with noise in the X-axis
# Add the data-points with noise in the X-axis
rug(jitter(X[,"x1"]))

# Plot density for x3
x3_density <- density(x3)
hist(x3,freq = FALSE,main = "Histogram and density plot of x3",xlab = "x3 Signal")
lines(x3_density,lwd=2,col="green")
# Add the data-points with noise in the X-axis
rug(jitter(X[,"x3"]))

# Plot density for x4
x4_density <- density(x4)
hist(x4,freq = FALSE,main = "Histogram and density plot of x4",xlab = "x4 Signal")
lines(x4_density,lwd=2,col="blue")
# Add the data-points with noise in the X-axis
rug(jitter(X[,"x4"]))

# Plot density for x5
x5_density <- density(x5)
hist(x5,freq = FALSE,main = "Histogram and density plot of x5",xlab = "x5 Signal")
lines(x5_density,lwd=2,col="purple")
# Add the data-points with noise in the X-axis
rug(jitter(X[,"x5"]))


# Plot density for Y
# Reset the plotting area to have 1 plot for the output variable
y_density <- density(y)
hist(y,freq = FALSE,main = "Histogram and density plot of Output Signal y",xlab = "y Signal")
lines(y_density,lwd=2,col="orange")
# Add the data-points with noise in the X-axis
rug(jitter(Y[,"y"]))

```


## Task 1.3: Creating scatter plot to identify correlation

```{r}
par(mfrow=c(2,2))

# Plotting input signal X1 against output signal Y
plot(X[,"x1"],Y,main = "Correlation betweeen X1 and Y signal", xlab = "X1 signal", ylab = "Output signal y")

# Plotting input signal X2 against output signal Y
plot(X[,"x3"],Y,main = "Correlation betweeen X3 and Y signal", xlab = "X3 signal", ylab = "Output signal y")

# Plotting input signal X3 against output signal Y
plot(X[,"x4"],Y,main = "Correlation betweeen X4 and Y signal", xlab = "X4 signal", ylab = "Output signal y")

# Plotting input signal X4 against output signal Y
plot(X[,"x5"],Y,main = "Correlation betweeen X5 and Y signal", xlab = "X5 signal", ylab = "Output signal y")
```


<!--Task 2: Begin-->
<!-- Normalize data -->

```{r}

# Apply Z-score normalization
normalize_zscore <- function(x) {
  return((x - mean(x)) / sd(x))
}

# Normalize all columns
data <- as.data.frame(lapply(rawData, normalize_zscore))

head(data)

```


## Assigning normalized data to X and Y
```{r}
X <- as.matrix(data[, 1:4]) # Select the first four columns
colnames(X)<-c("x1","x3","x4","x5")

Y <- as.matrix(rawData[, 5])   # Select the fifth x2 column (last one from non-normalized dataset)
colnames(Y)<-c("y")

# Accessing input and output variables from matrix
x1 <- X[,1]
x3 <- X[,2]
x4 <- X[,3]
x5 <- X[,4]
y <- Y[,1]
```


# Task 2
## Calculating ones for binding the data
```{r}
ones = matrix(1 , length(X)/4,1)
ones
```


# Task 2.1
# Calculating thetahat of each candidate model
```{r}
# For Model 1
#Binding data from equation of Model 1.
X_model1<-cbind(ones, X[,"x4"], X[,"x3"]^2)
# X_model1
#Calculating thetahat of Model 1
Model1_thetahat=solve(t(X_model1) %*% X_model1) %*% t(X_model1) %*% Y
Model1_thetahat


#For Model 2
#Binding data from equation of Model 2.
X_model2<-cbind(ones,X[,"x4"], X[,"x3"]^2, X[,"x5"])
# X_model2
#Calculating thetahat of Model 2
Model2_thetahat=solve(t(X_model2) %*% X_model2) %*% t(X_model2) %*% Y
Model2_thetahat


#Model 3
#Binding data from equation of Model 3.
X_model3<-cbind(X[,"x3"], X[,"x4"], X[,"x5"]^3)
# X_model3
#Calculating thetahat of Model 3
Model3_thetahat=solve(t(X_model3) %*% X_model3) %*% t(X_model3) %*% Y
Model3_thetahat


#For Model 4
#Binding data from equation of Model 4.
X_model4<-cbind(ones,(X[,"x4"]), X[,"x3"]^2, X[,"x5"]^3)
# X_model4
# #Calculating thetahat of Model 4
Model4_thetahat=solve(t(X_model4) %*% X_model4) %*% t(X_model4) %*% Y
Model4_thetahat


# for Model 5
#Binding data from equation of Model 5.
X_model5<-cbind(ones, X[,"x4"], X[,"x1"]^2, X[,"x3"]^2)
# X_model5
#Calculating thetahat of model 1
Model5_thetahat=solve(t(X_model5) %*% X_model5) %*% t(X_model5) %*% Y
Model5_thetahat

```

## Printing value of thetahat of each model
```{r}
# Model 1
# Model1_thetahat
cat("Model 1 thetahat: ")
c(Model1_thetahat)
#Model 2
# Model2_thetahat
cat("Model 2 thetahat: ")
c(Model2_thetahat)
#Model 3
# Model3_thetahat
cat("Model 3 thetahat: ")
c(Model3_thetahat)

#Model 4
# Model4_thetahat
cat("Model 4 thetahat: ")
c(Model4_thetahat)

#Model 5
# Model5_thetahat
cat("Model 5 thetahat: ")
c(Model5_thetahat)
```



## Task 2.2
## Calculating Y-hat and RSS for each model
```{r}
#Calculating Y-hat and RSS Model 1
Y_hat_model1 = X_model1 %*% Model1_thetahat
Y_hat_model1
#Calculating RSS
RSS_Model_1=sum((Y-Y_hat_model1)^2)
RSS_Model_1

# Calculating Y-hat and RSS of Model 2
Y_hat_model2 = X_model2 %*% Model2_thetahat
Y_hat_model2
#Calculating RSS
RSS_Model_2=sum((Y-Y_hat_model2)^2)
RSS_Model_2

# Calculating Y-hat and RSS of Model 3
Y_hat_model3 = X_model3 %*% Model3_thetahat
Y_hat_model3
#Calculating RSS
RSS_Model_3=sum((Y-Y_hat_model3)^2)
RSS_Model_3
 
# Calculating Y-hat and RSS of Model 4
Y_hat_model4 <- X_model4 %*% Model4_thetahat
head(Y_hat_model4)
Y_hat_model4
#Calculating RSS
RSS_Model_4=sum((Y-Y_hat_model4)^2)
RSS_Model_4

# Calculating Y-hat and RSS of Model 5
Y_hat_model5 = X_model5 %*% Model5_thetahat
Y_hat_model5
#Calculating RSS
RSS_Model_5=sum((Y-Y_hat_model5)^2)
RSS_Model_5
```


### Printing RSS values
```{r}
model1 <- c(RSS_Model_1)
model2 <- c(RSS_Model_2)
model3 <- c(RSS_Model_3)
model4 <- c(RSS_Model_4)
model5 <- c(RSS_Model_5)

dfRSS <- data.frame("Model1" = model1, "Model2"=model2, "Model3"=model3, "Model4"=model4, "Model5"=model5, stringsAsFactors = FALSE)
dfRSS
```


## Task 2.3 Calculating likelihood and Variance of each model
```{r}
N=length(Y)

#Calculating the Variance of Model 1
Variance_model1=RSS_Model_1/(N-1)
Variance_model1

#Calculating the log-likelihood of Model 1
likehood_Model_1=
  -(N/2)*(log(2*pi))-(N/2)*(log(Variance_model1))-(1/(2*Variance_model1))*RSS_Model_1
likehood_Model_1

#Calculating Variance and log-likelihood of Model 2
Variance_model2=RSS_Model_2/(N-1)
Variance_model2
likehood_Model_2=
  -(N/2)*(log(2*pi))-(N/2)*(log(Variance_model2))-(1/(2*Variance_model2))*RSS_Model_2
likehood_Model_2


#Calculating Variance and log-likelihood of Model 3
Variance_model3=RSS_Model_3/(N-1)
Variance_model3
likehood_Model_3=
  -(N/2)*(log(2*pi))-(N/2)*(log(Variance_model3))-(1/(2*Variance_model3))*RSS_Model_3
likehood_Model_3

#Calculating Variance and log-likelihood of Model 4
Variance_model4=RSS_Model_4/(N-1)
Variance_model4
likehood_Model_4=
  -(N/2)*(log(2*pi))-(N/2)*(log(Variance_model4))-(1/(2*Variance_model4))*RSS_Model_4
likehood_Model_4

#Calculating Variance and log-likelihood of Model 5
Variance_model5=RSS_Model_5/(N-1)
Variance_model5
likehood_Model_5=
  -(N/2)*(log(2*pi))-(N/2)*(log(Variance_model5))-(1/(2*Variance_model5))*RSS_Model_5
likehood_Model_5
```

### Printing variance  values
```{r}
model1 <- c(Variance_model1)
model2 <- c(Variance_model2)
model3 <- c(Variance_model3)
model4 <- c(Variance_model4)
model5 <- c(Variance_model5)

dfVariance <- data.frame(model1, model2,model3,model4,model5)
dfVariance
```


### Printing  likelihood values
```{r}
model1 <- c(likehood_Model_1)
model2 <- c(likehood_Model_2)
model3 <- c(likehood_Model_3)
model4 <- c(likehood_Model_4)
model5 <- c(likehood_Model_5)

dfLikelihood <- data.frame(model1, model2,model3,model4,model5)
dfLikelihood
```


## Task 2.4 
### Calculating AIC And BIC of each model
```{r}
# Calculating AIC and BIC of model 1
K_model1<-length(Model1_thetahat)
K_model1
AIC_model1=2*K_model1-2*likehood_Model_1
AIC_model1
BIC_model1=K_model1*log(N)-2*likehood_Model_1
BIC_model1

## thetahat of model 2
K_model2<-length(Model2_thetahat)
K_model2
##Calculating AIC and BIC of model 2
AIC_model2=2*K_model2-2*likehood_Model_2
AIC_model2
BIC_model2=K_model2*log(N)-2*likehood_Model_2
BIC_model2

## thetahat of model 3
K_model3<-length(Model3_thetahat)
K_model3
##Calculating AIC and BIC of model 3
AIC_model3=2*K_model3-2*likehood_Model_3
AIC_model3
BIC_model3=K_model3*log(N)-2*likehood_Model_3
BIC_model3

## thetahat of model 4
K_model4<-length(Model4_thetahat)
K_model4
##Calculating AIC and BIC of model 4
AIC_model4=2*K_model4-2*likehood_Model_4
AIC_model4
BIC_model4=K_model4*log(N)-2*likehood_Model_4
BIC_model4

## thetahat of model 5
K_model5<-length(Model5_thetahat)
K_model5
##Calculating AIC and BIC of model 5
AIC_model5=2*K_model5-2*likehood_Model_5
AIC_model5
BIC_model5=K_model5*log(N)-2*likehood_Model_5
BIC_model5
```

### Printing  K values

```{r}
model1 <- c(K_model1)
model2 <- c(K_model2)
model3 <- c(K_model3)
model4 <- c(K_model4)
model5 <- c(K_model5)

dfK <- data.frame(model1, model2,model3,model4,model5)
dfK
```

### printing AIC values
```{r}
model1 <- c(AIC_model1)
model2 <- c(AIC_model2)
model3 <- c(AIC_model3)
model4 <- c(AIC_model4)
model5 <- c(AIC_model5)

dfAIC <- data.frame(model1, model2,model3,model4,model5)
dfAIC
```


### Printing BIC values
```{r}
model1 <- c(BIC_model1)
model2 <- c(BIC_model2)
model3 <- c(BIC_model3)
model4 <- c(BIC_model4)
model5 <- c(BIC_model5)

dfBIC <- data.frame(model1, model2,model3,model4,model5)
dfBIC
```


## Task 2.5 Calculating error plotting normal/gaussian distibution of each plot
```{r}
par(mfrow=c(1,1))

## Error of model1
model1_error <- Y-Y_hat_model1
model1_error

## Plotting the graph QQplot and QQ line of model 1
qqnorm(model1_error, col = "darkblue",main = "QQ plot of model 1")
qqline(model1_error, col = "brown",lwd=1)


## Error of model2
model2_error <- Y-Y_hat_model2 # error of model 2
## Plotting QQplot and QQ line of model 2
qqnorm(model2_error, col = "darkblue",main = "QQ plot of model 2")
qqline(model2_error, col = "brown")


## Error of model3
model3_error <- Y- Y_hat_model3
## Plotting QQplot and QQ line of model 3
qqnorm(model3_error, col = "darkblue",main = "QQ plot of model 3")
qqline(model3_error, col = "brown")

## Error of model4
model4_error <- Y-Y_hat_model4
## Plotting QQplot and QQ line of model 4
qqnorm(model4_error, col = "darkblue",main = "QQ plot of model 4")
qqline(model4_error, col = "brown")

## Error of model5
model5_error <- Y- Y_hat_model5
## Plotting QQplot and QQ line of model 5
qqnorm(model5_error, col = "darkblue",main = "QQ plot of model 5")
qqline(model5_error, col = "brown")
```


## Task 2.7 splitting data into training and testing dataset and calculating estamation based on training dataset
##also plotting normal distribution graph of training data
```{r}
## Spliting the dataset y into  Traning and testing data set.
set.seed(253)
split_Y<-initial_split(data = as.data.frame(Y),prop=.7)
## Traning splitted Y dataset 
Y_training_set<-training(split_Y)
Y_testing_set<-as.matrix(testing(split_Y))
## Testing splitted Y dataset 
Y_training_data<-as.matrix(Y_training_set)

## Spliting the dataset of X into  Traning and testing data set.
split_X<-initial_split(data = as.data.frame(X),prop=.7)
## Traning splitted X dataset
X_training_set<-training(split_X)
## Testing splitted X dataset 
X_testing_set<-as.matrix(testing(split_X))
X_testing_data<-as.matrix(X_testing_set)
X_training_data<-as.matrix(X_training_set)

### Estimating model parameters using Traning set
traning_ones=matrix(1 , length(X_training_set$x1),1)
# selected model 3 and using equation of model 3
X_traning_model<-cbind(traning_ones,X_training_set[,"x3"],(X_training_set[,"x4"]),(X_training_set[,"x5"])^3)
traning_thetahat=solve(t(X_traning_model) %*% X_traning_model) %*% t(X_traning_model) %*%  Y_training_data
  
### Model out/Prediction
Y_testing_hat = X_testing_data %*% traning_thetahat
Y_testing_hat
RSS_testing=sum((Y_testing_set-Y_testing_hat)^2)
RSS_testing
t.test(Y_training_data, mu=500, alternative="two.sided", conf.level=0.95)
C_I1=454.0084
C_I2=454.8277
p2 <- plot(density(Y_training_data), col="blue", lwd=2,
           main="Distribution of Traning Data")
abline(v=C_I1,col="brown", lty=2)
abline(v=C_I2,col="brown", lty=2)

thetaHat_training =solve(t(X_training_data) %*% X_training_data) %*% t(X_training_data) %*%Y_training_data
thetaHat_training
length(thetaHat_training)
dis_test=density(Y_training_data)
plot((dis_test))
plot(dis_test,main = "Density plot of Y Signal")

### Calculating Confidential interval
z=1.96 ##(95%) Confidential interval
error=((Y_testing_set-Y_testing_hat))
n_len=length(Y_testing_hat)
C_I_1= z * sqrt( (error * (1-error) ) / n_len)
C_I_1
error
C_I_2= z * sqrt( (error * (1+error)) / n_len)
C_I_2            

```


## Task 3
```{r}
## Model 2 will be used, parameter are selected and kept constant.
f_value=0
s_value=0
arr_1=0
arr_2=0
#values from thetahat
thetabias <- abs(Model2_thetahat[1]) #selected parameter
thetaone <- abs(Model2_thetahat[2]) # selected parameter
thetatwo <- abs(Model2_thetahat[3]) # constant value
thetathree <- abs(Model2_thetahat[4]) # constant value


Epison <- RSS_Model_2 * 2 ## fixing value of epison
num <- 100 #number of iteration
##Calculating Y-hat for performing rejection ABC
range1 <- runif(1, thetabias * 0.8, thetabias * 1.2)
range1

range2 <- runif(1, thetaone * 0.8, thetaone * 1.2)
range2
counter <- 0
for (i in 1:num) {
  range1 <- runif(1, thetabias * 0.8, thetabias * 1.2) # calculating the range
  range2 <- runif(1, thetaone * 0.8, thetaone * 1.2)
  
  New_thetahat <- matrix(c(range1,range2,thetatwo,thetathree))
  New_Y_Hat <- X_model2 %*% New_thetahat ## calculating new Y-hat
  new_RSS <- sum((Y-New_Y_Hat)^2)
  new_RSS
  if (new_RSS > Epison){
    arr_1[i] <- range1
    arr_2[i] <- range2
    counter = counter+1
    f_value <- matrix(arr_1)
    s_value <- matrix(arr_2)
  }
}
hist(f_value)
hist(s_value)

###ploting Joint and Marginal Posterior Distribution of the graph
plot(f_value,s_value, col = c("brown", "blue"), main = "Joint and Marginal Posterior Distribution")
par(mfrow=c(1,1))
```







